Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-11-21 paper/paper.tex
move text to PLoS template

\documentclass[9pt,lineno]{template}

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage[export]{adjustbox}
\usepackage{float}
\DeclareSIUnit\Molar{M}

\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
\newcommand{\abrcomment}[1]{\emph{\color{blue} [#1]}}

\renewcommand{\floatpagefraction}{.9}

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{Supplementary File}

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{Supplementary Figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}

\author[1]{Alistair B. Russell}
\author[1]{Jacob R. Kowalsky}
\author[1,2,3*]{Jesse D. Bloom}
\affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States}
\affil[2]{Department of Genome Sciences, University of Washington, Seattle, United States}
\affil[3]{Howard Hughes Medical Institute, Fred Hutchinson Cancer Research Center, Seattle, United States}

\corr{jbloom@fredhutch.org}{JDB}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
\end{abstract}

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.

@@

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\citep{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Alistair B. Russell\textsuperscript{1},
Jacob R. Kowalsky\textsuperscript{2},
Jesse D. Bloom\textsuperscript{1,2,3*}
\\
\bigskip
\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
\\
\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
\\
\textbf{3} Howard Hughes Medical Institute, Seattle, United States
\\
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* jbloom@fredhutch.org

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
{\color{red} add this}

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.

@@

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\cite[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\cite[CCSs;][]{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
